---
title: "Qiime2 commands for 2022-2024 MOCH samples"
author: "Gemma Clucas"
date: "2023-05-08"
output: github_document
---

```{r setup, include=FALSE}
options(scipen=999)
knitr::opts_chunk$set(echo = TRUE, digits = 1)
knitr::opts_chunk$set(save_output = FALSE) # prevent outputs being over-written on knitting
library(tidyverse)
library(knitr)
library(qiime2R)
```

## 1. Import the data
Just one plate for the 2022 samples.
```
cd /Users/gemmaclucas/GitHub/Fecal_metabarcoding/Mountain_Chickadee_diets/2022
conda activate qiime2-2021.4

qiime tools import\
  --type 'SampleData[PairedEndSequencesWithQuality]'\
  --input-path /Volumes/Data_SS1/ANML/LaurenWhitenack_MountainChickadees/Plate34/reads  \
  --input-format CasavaOneEightSingleLanePerSampleDirFmt\
  --output-path demux_plate1.qza
  
for K in {1..1}; do
  qiime demux summarize \
    --i-data demux_Plate$K.qza \
    --o-visualization demux_Plate$K.qzv
done
```

### 2023 samples
The data for both plates of samples from 2023 was combined into one sequencing run at UNH, so it's all in one folder.
```
/Users/gc547/Dropbox/GitHub_copied/Fecal_metabarcoding/Mountain_Chickadee_diets/2023
conda activate qiime2-amplicon-2024.10

qiime tools import\
  --type 'SampleData[PairedEndSequencesWithQuality]'\
  --input-path /Volumes/Data_SS1/ANML/LaurenWhitenack_MountainChickadees/Plate1and2_2023/reads  \
  --input-format CasavaOneEightSingleLanePerSampleDirFmt\
  --output-path demux_plate1.qza
  
for K in {1..1}; do
  qiime demux summarize \
    --i-data demux_Plate$K.qza \
    --o-visualization demux_Plate$K.qzv
done
```

Blanks look very clean. Lots of variability in read numbers between samples though.

### 2024 samples
Samples are mostly on MOCH-2024-1 with a handful on MOCH-2024-2 alongside woodpecker samples.
```
/Users/gc547/Dropbox/GitHub_copied/Fecal_metabarcoding/Mountain_Chickadee_diets/2024
conda activate qiime2-amplicon-2024.10

qiime tools import\
  --type 'SampleData[PairedEndSequencesWithQuality]'\
  --input-path /Volumes/Data_SS1/ANML/LaurenWhitenack_MountainChickadees/2024_samples/MOCH-2024-1/MOCH_reads  \
  --input-format CasavaOneEightSingleLanePerSampleDirFmt\
  --output-path demux_plate1.qza
  
qiime tools import\
  --type 'SampleData[PairedEndSequencesWithQuality]'\
  --input-path /Volumes/Data_SS1/ANML/LaurenWhitenack_MountainChickadees/2024_samples/MOCH-2024-2/MOCH_reads  \
  --input-format CasavaOneEightSingleLanePerSampleDirFmt\
  --output-path demux_plate2.qza
  
for K in {1..2}; do
  qiime demux summarize \
    --i-data demux_Plate$K.qza \
    --o-visualization demux_Plate$K.qzv
done
```


## 2. Trim primers using cutadapt

The primer sequences are:   

ANML_F GGTCAACAAATCATAAAGATATTGG  
ANML_R GGWACTAATCAATTTCCAAATCC 

### Trim 3' ends first

At the 3' end of the read, the primer will have been read through after reading the ANML region. I need to be looking for the reverse complement of the reverse primer in read 1 (—p-adapter-f) and the reverse complement of the forward primer in R2 (—p-adapter-r).

R primer reverse complement: GGATTTGGAAATTGATTAGTWCC  
F primer reverse complement: CCAATATCTTTATGATTTGTTGACC  
```
for K in {1..1}; do
  qiime cutadapt trim-paired \
    --i-demultiplexed-sequences demux_Plate$K.qza \
    --p-adapter-f GGATTTGGAAATTGATTAGTWCC \
    --p-adapter-r CCAATATCTTTATGATTTGTTGACC \
    --o-trimmed-sequences trimd_Plate$K.qza \
    --verbose > cutadapt_out_Plate$K.txt
done
```
To see how much data passed the filter:
```
grep "Total written (filtered):" cutadapt_out_Plate1.txt 
```
About 82% seems to have passed the filters for each sample, which is the same as in the BTBW samples we sequenced.

### Trim 5' ends of reads

All R1 should begin with the forward primer: GGTCAACAAATCATAAAGATATTGG (25 bases).  
All R2 should begin with the reverse primer: GGWACTAATCAATTTCCAAATCC (23 bases).  

Trim these with the following commands:
```
for K in {1..1}; do
  qiime cutadapt trim-paired \
    --i-demultiplexed-sequences trimd_Plate$K.qza \
    --p-front-f GGTCAACAAATCATAAAGATATTGG \
    --p-front-r GGWACTAATCAATTTCCAAATCC  \
    --o-trimmed-sequences trimd2_Plate$K.qza \
    --verbose > cutadapt_out2_Plate$K.txt
done
```
To see how much data passed the filter for each sample:
```
grep "Total written (filtered):" cutadapt_out2_Plate1.txt 
```

88% this time - normal.

### 2023 Samples

Exact same code as above. 82% kept after the first trim, 88% after the second, as before.

### 2024 Samples

Exact same code as above but with the for loop doing two plates. 80-82% kept after the first trim, 88% after the second, as before.


## 3. Denoise with Dada2
Using the same settings that seemed to work well for the BTBW samples. Truncating the reads to 130bp avoided reads being lost to the low-quality filter (because read quality dropped after about 160bp as we got to the end of the reads) and given that the amplicons are usually about 185bp in length, we would actually expect an overlap of 75bp on average, so this is plenty to ensure proper merging.

```
for K in {1..1}; do
  qiime dada2 denoise-paired \
    --i-demultiplexed-seqs trimd2_Plate$K.qza \
    --p-trunc-len-f 130 \
    --p-trunc-len-r 130 \
    --p-trim-left-f 0 \
    --p-trim-left-r 0 \
    --p-min-overlap 50 \
    --p-n-threads 8 \
    --o-representative-sequences rep-seqs_Plate$K \
    --o-table table_Plate$K \
    --o-denoising-stats denoise_Plate$K
done


for K in {1..1}; do  
  qiime metadata tabulate\
    --m-input-file denoise_Plate$K.qza\
    --o-visualization denoise_Plate$K.qzv
done
```

To view the rep-seqs and tables:
```
for K in {1..1}; do  
  qiime feature-table tabulate-seqs \
    --i-data rep-seqs_Plate$K.qza \
    --o-visualization rep-seqs_Plate$K
done

for K in {1..1}; do  
  qiime feature-table summarize \
    --i-table table_Plate$K.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization table_Plate$K
done
```

### 2023 Samples

Exact same code as above. Still some reads in one of the extraction blanks, but the number is very low compared to the samples.

### 2024 Samples

Exact same code as above but with the for loop going through both plates. 

Since there are two plates of data, we need to merge before assigning taxonomy
```
qiime feature-table merge \
  --i-tables table_Plate1.qza \
  --i-tables table_Plate2.qza \
  --p-overlap-method sum \
  --o-merged-table merged-table.qza

qiime feature-table summarize \
    --i-table merged-table.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization merged-table
    
qiime feature-table merge-seqs \
  --i-data rep-seqs_Plate1.qza \
  --i-data rep-seqs_Plate2.qza \
  --o-merged-data merged_rep-seqs.qza

qiime feature-table tabulate-seqs \
  --i-data merged_rep-seqs.qza \
  --o-visualization merged_rep-seqs.qzv
```



## 4. Assign taxonomy

Using the same COI database and classifier that Devon O'Rourke trained. Details (here)[https://forum.qiime2.org/t/building-a-coi-database-from-bold-references/16129]. 

I downloaded the database sequences (bold_anml_seqs.qza), the taxonomy strings that go with each sequence (bold_anml_taxa.qza) and the trained classifier (bold_anml_classifier.qza) into this folder. Note this only runs with the 2019 version of QIIME, which is what he must have been running when he made the classifier.

```
conda activate qiime2-2019.10

qiime feature-classifier classify-sklearn \
  --i-classifier bold_anml_classifier.qza \
  --i-reads rep-seqs_Plate1.qza \
  --o-classification taxonomy.qza
  
qiime metadata tabulate \
  --m-input-file taxonomy.qza \
  --o-visualization taxonomy.qzv
```

### 2023 Samples

I couldn't use the original classifier because I can't run the 2019 version of QIIME on this new M3 laptop (and it's not available on the cluster either). So I trained a classifier using the same database of sequences that Devon had trained the previous classifier on (bold_anml_seqs.qza and bold_anml_taxa.qza), so that they are comparable, and this works now. 

```
qiime feature-classifier classify-sklearn \
  --i-classifier classifier.qza \
  --i-reads rep-seqs_Plate1.qza \
  --o-classification taxonomy.qza
  
qiime metadata tabulate \
  --m-input-file taxonomy.qza \
  --o-visualization taxonomy.qzv
```

### 2024 Samples
Using the classifier from 2023.
```
qiime feature-classifier classify-sklearn \
  --i-classifier ../2023/classifier.qza \
  --i-reads merged_rep-seqs.qza \
  --o-classification taxonomy.qza
  
qiime metadata tabulate \
  --m-input-file taxonomy.qza \
  --o-visualization taxonomy.qzv
```


## 5. Make some barplots

Just to see what's in the samples.

```
qiime taxa barplot \
  --i-table table_Plate1.qza \
  --i-taxonomy taxonomy.qza \
  --m-metadata-file metadata.txt \
  --o-visualization barplot_before_filtering.qzv
```

### 2023 Samples

Same as above. Massive diversity of insects. Also some rodent and bear DNA!

### 2024 Samples

```
qiime taxa barplot \
  --i-table merged-table.qza \
  --i-taxonomy taxonomy.qza \
  --m-metadata-file metadata.txt \
  --o-visualization barplot_before_filtering.qzv
```
Massive diversity of insects again. Bear, chipmunk, and red squirrel come up too.

## 6. Remove non-arthropod reads and drop mock community and blanks

```
qiime taxa filter-table \
  --i-table table_Plate1.qza \
  --i-taxonomy taxonomy.qza \
  --p-include Arthropoda \
  --o-filtered-table table_Arthropoda.qza
  
qiime feature-table summarize \
    --i-table table_Arthropoda.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization table_Arthropoda
```
After removing non-arthropod reads, one of the blanks (34-3) had some arthropod reads, but it's only 14 reads so it is very minimal.

I am also going to remove reads that were only classified to Arthropoda and not beyond, since they are not giving us any useful information. So that means I can use the `--p-include` flag to include anything that also has a class designation, since they all have `Arthropoda;c_` in the taxonomy string.

```
qiime taxa filter-table \
  --i-table table_Arthropoda.qza \
  --i-taxonomy taxonomy.qza \
  --p-include "Arthropoda;c_" \
  --o-filtered-table table_ArthropodaC.qza
  
qiime feature-table summarize \
    --i-table table_ArthropodaC.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization table_ArthropodaC
```

Drop the blanks and mock community and then filter the rep-seqs for just the species that are in the table_ArthropodaC, so that I can send that and sequence counts to Lauren, so that she has the unrarefied data.

```
qiime feature-table filter-samples \
  --i-table table_ArthropodaC.qza \
  --m-metadata-file metadata.txt \
  --p-where "Species='MOCH'" \
  --o-filtered-table table_ArthropodaC_MOCH.qza

qiime feature-table filter-seqs \
  --i-data rep-seqs_Plate1.qza \
  --i-table table_ArthropodaC_MOCH.qza \
  --o-filtered-data rep-seqs_ArthropodaC_MOCH.qza

qiime metadata tabulate \
  --m-input-file rep-seqs_ArthropodaC_MOCH.qza \
  --m-input-file taxonomy.qza \
  --o-visualization sequence_taxonomy_2022ArthropodaMOCH.qzv
```
Also make a barplot to download the final dataset as a csv.
```
qiime taxa barplot \
  --i-table table_ArthropodaC_MOCH.qza \
  --m-metadata-file metadata.txt \
  --i-taxonomy taxonomy.qza \
  --o-visualization barplot_ArthropodaC_MOCH
```  

### 2023 Samples

Same as above but going straight to the `--p-include "Arthropoda;c_"` step as the previous step (filtering for just arthropods) was redundant with the addition of this one.
BLANK-MOCH-1-1 has 4 reads, while BLANK-MOCH-1-2 has 134 reads, so this is nothing compared to the average depth of the samples.

```
qiime taxa filter-table \
  --i-table table_Plate1.qza \
  --i-taxonomy taxonomy.qza \
  --p-include "Arthropoda;c_" \
  --o-filtered-table table_ArthropodaC.qza
  
qiime feature-table summarize \
    --i-table table_ArthropodaC.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization table_ArthropodaC
```

### 2024 Samples
Same as 2023 code for selecting Arthropods with a class identified, but done on the merged table.
```
qiime taxa filter-table \
  --i-table merged-table.qza \
  --i-taxonomy taxonomy.qza \
  --p-include "Arthropoda;c_" \
  --o-filtered-table table_ArthropodaC.qza
  
qiime feature-table summarize \
    --i-table table_ArthropodaC.qza \
    --m-sample-metadata-file metadata.txt \
    --o-visualization table_ArthropodaC

```

Calculate depth of samples vs blanks
```{r}
source("/Users/gc547/Dropbox/GitHub_copied/Fecal_metabarcoding/Scripts/read_depth_analysis.R")

results <- analyze_read_depths(
  feature_table_path = "2024/table_ArthropodaC.qza",
  taxonomy_path = "2024/taxonomy.qza",
  metadata_path = "2024/metadata.txt",
  output_folder = "2024/", # 
  save_output = TRUE
)
```
The extraction blanks have 0.06% the depth of the samples, while the PCR blanks have 0.24% the depth of the samples on average, so we are safe to drop them and proceed.

Drop the blanks and mock community and then filter the rep-seqs for just the species that are in the table_ArthropodaC.qza.

```
qiime feature-table filter-samples \
  --i-table table_ArthropodaC.qza \
  --m-metadata-file metadata.txt \
  --p-where "Species='MOCH'" \
  --o-filtered-table table_ArthropodaC_MOCH.qza

qiime feature-table filter-seqs \
  --i-data merged_rep-seqs.qza \
  --i-table table_ArthropodaC_MOCH.qza \
  --o-filtered-data rep-seqs_ArthropodaC_MOCH.qza

qiime metadata tabulate \
  --m-input-file rep-seqs_ArthropodaC_MOCH.qza \
  --m-input-file taxonomy.qza \
  --o-visualization sequence_taxonomy_2024ArthropodaMOCH.qzv
```
Also make a barplot.
```
qiime taxa barplot \
  --i-table table_ArthropodaC_MOCH.qza \
  --m-metadata-file metadata.txt \
  --i-taxonomy taxonomy.qza \
  --o-visualization barplot_ArthropodaC_MOCH
``` 
Have not sent anything to Lauren yet.
